---
title: "Case study: copula dependence between temperature and ozone"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: copula dependence between temperature and ozone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

A correlation coefficient answers one question: how much do two variables
move together on average. It is silent on a different question that often
matters more: do they move together in the extremes. Two variables can
share a modest correlation and still spike together in their upper tails,
or be nearly independent in the body and coupled only when both are large.
A copula separates that dependence structure from the marginal shapes and
lets it be read directly. This case study fits a bivariate copula to a
real pair, the daily maximum temperature and ozone concentration at a New
York monitoring station (the `airquality` data), and asks not only how
strong the dependence is but where in the joint distribution it lives. The
answer overturns what the raw scatter appears to show, which is the
lesson.

```{r}
library(gpumetropolis)
aq <- airquality[stats::complete.cases(airquality[, c("Temp", "Ozone")]), ]
x <- aq$Temp
y <- aq$Ozone
plot(x, y, pch = 16, col = grDevices::adjustcolor("black", 0.5),
     xlab = "temperature (F)", ylab = "ozone (ppb)",
     main = "Ozone against temperature, New York, 1973")
```

The scatter suggests both a positive trend and a fan: the hottest days
carry not just higher ozone but a wider and higher spread. The eye reads
that fan as a hint of upper-tail dependence, extremes travelling together
at the top. That reading is a hypothesis, and the copula is what tests it;
the answer below is not the one the scatter suggests, which is the point
of the exercise.

## 1. Pseudo-observations, the dependence stripped of the margins

The copula lives on the unit square, and the data enter it as
pseudo-observations: each value is replaced by its rank divided by
`n + 1`, which maps the marginal to a uniform and leaves only the
dependence. `gpum_copula()` does this internally; here is the picture it
works from.

```{r}
u <- rank(x) / (length(x) + 1)
v <- rank(y) / (length(y) + 1)
plot(u, v, pch = 16, col = grDevices::adjustcolor("black", 0.5),
     xlab = "u = rank(temperature)", ylab = "v = rank(ozone)",
     main = "Pseudo-observations on the unit square")
```

The points look denser toward the upper-right corner than the lower-left,
which would be upper-tail dependence. Whether that apparent asymmetry is
real dependence or an artefact of the eye is what the formal comparison
settles.

## 2. Fitting with automatic family selection

Rather than assume a family, `family = "auto"` fits all four candidates
and selects the one preferred by predictive comparison (WAIC on the
copula density, which is a proper density on the unit square). The choice
of family is itself a hypothesis about tail behaviour, so letting the data
choose is part of the answer.

```{r}
fit <- gpum_copula(data.frame(temp = x, ozone = y), family = "auto",
                   n_iter = 12000, n_chains = 8, seed = 1, backend = "cpu")
fit
```

The printed object is the technical diagnostic: the selected family, the
dependence parameter and Kendall's tau with their credible intervals, the
tail-dependence coefficients, and a verdict line that reports convergence
(R-hat) and goodness of fit (whether the empirical Kendall tau falls
inside the model's posterior band). The comparison table ranks the four
families by WAIC.

The data select Frank, not Gumbel. Frank is a symmetric, tail-independent
copula: strong overall dependence (Kendall's tau near 0.56) with no
concentration in either tail. The upper-tail hint the scatter suggested
did not survive the comparison; the Gumbel family, which would impose
upper-tail dependence, ranks below Frank on WAIC. The fan in the raw
scatter was heteroscedasticity of the ozone margin, higher variance at
high temperature, not a coupling of the extremes. This is the discipline
the copula enforces: the dependence structure is read after the marginal
shapes are removed by the ranks, and the eye's reading of the raw scatter,
which mixes the two, is overruled.

## 3. Reading the dependence

Kendall's tau is the family-free summary, comparable across copulas; the
tail-dependence coefficients are the extreme-co-movement probabilities.

```{r}
kendall_tau(fit)
tail_dependence(fit)
```

Both tail-dependence coefficients are zero, the signature of the Frank
family: temperature and ozone are strongly associated through the body of
the distribution, but their extremes are asymptotically independent.
Knowing that today is among the hottest on record does not, in the tail
limit, make a record ozone reading more likely than the overall
association already implies. That is a materially different risk statement
from the upper-tail dependence the scatter suggested, and it is the copula,
not the eye, that is entitled to make it.

## 4. Visual inspection

The `plot()` method is the visual companion of the printed diagnostic:
the fitted density contours over the pseudo-observations, the posteriors
of the parameter and of Kendall's tau against the empirical value, and the
observed-against-generated check on the unit square.

```{r}
plot(fit)
```

The Frank contours are symmetric about the anti-diagonal, no corner
favoured, the posterior of tau sits on the empirical mark, and the
generated blue cloud overlays the observed black one. That overlay is the
confirmation that a symmetric, tail-independent copula reproduces the
observed dependence; had the upper tail been genuinely coupled, the
generated cloud would fall short of the observed points in the upper-right
corner.

## 5. Against the classical estimator

The `copula` package fits the same families by maximum pseudo-likelihood.
The Bernstein-von Mises theorem says the Bayesian posterior mean and the
maximum-likelihood estimate must agree for a regular model at this sample
size; the check confirms the fit is not an artefact of the sampler.

```{r}
if (requireNamespace("copula", quietly = TRUE)) {
  cop <- switch(fit$family,
    gumbel = copula::gumbelCopula(), clayton = copula::claytonCopula(),
    frank = copula::frankCopula(), gaussian = copula::normalCopula())
  ml <- copula::fitCopula(cop, cbind(u, v), method = "mpl")
  cat(sprintf("family %s: posterior mean %.3f vs pseudo-MLE %.3f\n",
              fit$family, fit$param["mean"], as.numeric(coef(ml))))
  cat(sprintf("Kendall tau: posterior %.3f vs empirical %.3f\n",
              fit$tau["mean"], fit$tau_empirical))
}
```

## 6. Forcing the eye's hypothesis, and seeing it rejected

The scatter argued for upper-tail dependence, which is the Gumbel family.
Fitting Gumbel by force shows what imposing that hypothesis would claim,
and the comparison of step 2 already recorded that it fits worse. Read the
two side by side: Gumbel would assert a positive upper tail-dependence
coefficient the data do not support, at a worse WAIC, while agreeing with
Frank on the overall strength.

```{r}
fit_gumbel <- gpum_copula(data.frame(temp = x, ozone = y),
                          family = "gumbel", n_iter = 8000, n_chains = 8,
                          seed = 1, backend = "cpu")
rbind(
  selected_frank = c(tau = fit$tau["mean"],
                     tail_upper = fit$tail_upper["mean"]),
  forced_gumbel = c(tau = fit_gumbel$tau["mean"],
                    tail_upper = fit_gumbel$tail_upper["mean"])
)
```

The two families agree on the overall strength of dependence and disagree
on where it lives: Frank places none in the tails, Gumbel would place a
substantial share in the upper tail. Choosing the family is choosing a
hypothesis about the extremes, and the copula workflow makes that choice an
inference the data adjudicate rather than an assumption the analyst
imports.

## 7. Where this stops

The workflow here is bivariate. Three or more variables need the vine
decomposition of a later release, which factors a high-dimensional
dependence into a cascade of bivariate copulas, each of the kind fit
here. The marginal distributions were handled non-parametrically through
the ranks; the companion release on marginal auto-selection fits an
explicit distribution to each column so the two halves of Sklar's theorem,
margins and copula, are both modelled. And a caveat of method: heavy ties
in either variable blur the pseudo-observations and the copula's
identifiability (Genest and Neslehova 2007); the ozone and temperature
readings here are near-continuous, so the ranks are clean.

## References

- Genest, C., Neslehova, J. (2007). A primer on copulas for count data.
  *ASTIN Bulletin*, 37(2), 475-515.
- Nelsen, R. B. (2006). *An Introduction to Copulas*, 2nd ed. Springer.
- Sklar, A. (1959). Fonctions de repartition a n dimensions et leurs
  marges. *Publications de l'Institut de Statistique de l'Universite de
  Paris*, 8, 229-231.
