---
title: "Case study: automatic marginal selection and the full Sklar workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: automatic marginal selection and the full Sklar workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5)
```

Sklar's theorem factors a joint distribution into its marginals and a
copula. The copula vignette handled the dependence half with the rank
transform, which ignores the marginal shapes. This one handles the other
half: fitting an explicit parametric distribution to each column, chosen
automatically, and then joining the two halves. The result is a full
probabilistic model of the joint distribution, with named marginals and a
named copula, the foundation for the synthesis of a later release.

```{r}
library(gpumetropolis)
```

## 1. One column, many candidate shapes

`gpum_fit_catalog()` takes a numeric column, detects its support and its
modality, fits every eligible parametric family with the sampler, and
ranks them by predictive comparison. Take the waiting times of the Old
Faithful geyser, a positive and visibly non-Gaussian variable.

```{r}
y <- faithful$waiting
result <- gpum_fit_catalog(y, n_iter = 8000, n_chains = 8, seed = 1,
                           backend = "cpu")
result
```

The printed object is the technical diagnostic: the detected support and
modality, the selected family, a convergence and goodness-of-fit verdict
(R-hat and the Kolmogorov-Smirnov distance of the fitted CDF against the
empirical), and the full ranking. The waiting times are bimodal, the dip
test rejects unimodality, and the catalogue selects the two-component
Gaussian mixture. The mixture is initialised with its components at the
lower and upper data quartiles and its means carry a data-anchored prior,
so the components start separated on the two modes and the gradient sampler
mixes cleanly within each; the draws are then relabeled so the
lower-location component is reported first. Every family is fit in an
unconstrained parametrisation, so the sampler never proposes an invalid
scale or shape.

## 2. Visual inspection

The `plot()` method is the visual companion: the fitted density over the
histogram (observed against generated), the QQ plot of empirical against
fitted quantiles, and the ranking of the candidates.

```{r, fig.height = 3.2}
plot(result)
```

The mixture density should sit on both humps of the histogram, the QQ
points should track the diagonal, and the ranking bar should show the
mixture clearly ahead of the unimodal families that cannot represent two
modes.

## 3. The fitted CDF is the transform to the unit square

The winner carries a fitted cumulative distribution, the
probability-integral transform. Where the copula vignette used ranks,
here the explicit fitted CDF carries the column onto the unit interval,
and the two agree closely when the marginal is well chosen.

```{r}
u_fitted <- marginal_cdf(result, y)
u_rank <- rank(y) / (length(y) + 1)
plot(u_rank, u_fitted, pch = 16, cex = 0.6,
     col = grDevices::adjustcolor("black", 0.5),
     xlab = "rank transform", ylab = "fitted-CDF transform",
     main = "The two probability-integral transforms agree")
abline(0, 1, col = "firebrick", lwd = 1.5, lty = 2)
```

## 4. The full Sklar workflow on two columns

With a fitted marginal per column, the dependence is modelled on the
fitted-CDF transforms rather than the ranks, so the marginals and the
copula together form one explicit model. Take the two variables of the
`airquality` temperature and ozone pair, each given its own automatically
selected marginal, then joined by an automatically selected copula.

```{r}
aq <- airquality[stats::complete.cases(airquality[, c("Temp", "Ozone")]), ]
m_temp <- gpum_fit_catalog(aq$Temp, n_iter = 6000, n_chains = 8, seed = 1,
                           backend = "cpu")
m_ozone <- gpum_fit_catalog(aq$Ozone, n_iter = 6000, n_chains = 8, seed = 1,
                            backend = "cpu")
c(temperature = m_temp$best, ozone = m_ozone$best)
```

```{r}
# Transform each column by its fitted CDF, then fit the copula on the
# normal scores of those transforms.
clamp <- function(p) pmin(pmax(p, 1e-4), 1 - 1e-4)
u <- clamp(marginal_cdf(m_temp, aq$Temp))
v <- clamp(marginal_cdf(m_ozone, aq$Ozone))
cop <- gpum_copula(data.frame(u = qnorm(u), v = qnorm(v)),
                   family = "auto", n_iter = 8000, n_chains = 8, seed = 1,
                   backend = "cpu")
cop
```

The joint model now has three named pieces: the marginal of temperature,
the marginal of ozone, and the copula that binds them, each selected by
the data. This is the complete factorisation Sklar's theorem promises,
and it is what the synthesis release will sample from to generate
synthetic records.

## 5. Where this stops

The catalogue here fits one column at a time and covers the common
continuous families plus the two-component Gaussian mixture; the full
catalogue design lists more families and the discrete tier. The mixture
handles bimodality; higher-order multimodality needs the three-component
mixture, and a dependence structure that is itself a mixture is a copula
question, not a marginal one. The fitted CDF is a plug-in at the
posterior mean; a fully Bayesian transform would average the CDF over the
posterior draws, which the synthesis release will use where the extra
fidelity matters.
