---
title: "Case study: the Nile flow, Bayesian inference beyond i.i.d."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: the Nile flow, Bayesian inference beyond i.i.d.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5)
```

Every case study so far treated the rows of the data as independent. This
one drops that assumption in the two ways it can fail, on a real series:
the annual flow of the Nile at Aswan, 1871 to 1970, a dataset built into R.
The series is famous for two features that map exactly onto the regimes
this vignette teaches: consecutive years are correlated (dependence), and
the completion of the first Aswan dam around 1898 shifted the level of the
whole series (non-identical distribution).

```{r}
library(gpumetropolis)
y <- as.numeric(Nile)
year <- 1871:1970
plot(year, y, type = "l", lwd = 1.5,
     main = "Annual flow of the Nile at Aswan",
     xlab = "year", ylab = expression("flow (" * 10^8 ~ m^3 * ")"))
abline(v = 1898, lty = 2)
text(1900, max(y), "Aswan dam", adj = 0)
```

## 1. The three ways data stop being i.i.d.

The sampler evaluates a per-row log-density and sums the rows. Under
independence that sum is the log-likelihood, whether or not the rows share
one distribution; under dependence the same sum is still the exact
log-likelihood as long as each row is the *conditional* density of its
observation given the previous ones. That single idea organises the three
regimes.

1. **Independent, not identically distributed.** Each row has its own law
   through its covariates. Every regression in the earlier case studies is
   already here; the theory changes name (Lindeberg-Feller and local
   asymptotic normality instead of the i.i.d. central limit theorem), the
   code does not change at all.
2. **Identically distributed, not independent.** A stationary series with
   Markov dependence. The joint density factorises as
   $p(y_{1:n}) = p(y_{1:p}) \prod_{t > p} p(y_t \mid y_{t-1}, \dots, y_{t-p})$,
   so the engine's row sum computes the conditional log-likelihood when the
   rows carry the lagged values. `gpum_ts_model()` builds that design.
3. **Neither independent nor identically distributed.** Dependence and
   covariates at once: the conditional density of each row also moves with
   exogenous variables. The Nile with the dam indicator is exactly this.

The asymptotics that license the posterior summaries are worth stating
once, plainly. In regime 2 and 3 the score of the conditional
factorisation is a martingale difference sequence, so the martingale
central limit theorem (Hall and Heyde 1980) plays the role the classical
CLT plays for independent data; maximum likelihood for ergodic Markov
processes behaves as in Billingsley (1961); and the Bernstein-von Mises
theorem carries over to ergodic Markov processes (Borwanker, Kallianpur
and Prakasa Rao 1971), which is what lets the credible intervals below be
read the usual way. Stationarity and ergodicity are model assumptions;
no diagnostic below can prove them.

## 2. Regime 2: a stationary AR(1) for the flow

The AR(1) conditional density of one row given its lag is Gaussian, and
the whole model is three parameters: the level `mu`, the persistence
`phi`, the innovation scale on the log scale `ls`. The formula is the
conditional log-density of the current value `y` given `y_lag1`; the
normalising constant is included so the model can enter predictive
comparisons later.

```{r}
m_ar1 <- gpum_ts_model(
  ~ -0.5 * ((y - mu - phi * (y_lag1 - mu)) / exp(ls))^2 - ls -
    0.9189385332046727,
  params = c("mu", "phi", "ls"),
  series = list(y = y), order = 1,
  prior = ~ -0.5 * ((mu - 900) / 500)^2 - 0.5 * (phi / 2)^2 -
    0.5 * ((ls - 5) / 3)^2
)
m_ar1
```

The conditional factorisation gives 99 rows from 100 observations: the
first year is conditioned on, the standard treatment of the initial state.
The fit is a regular `gpu_metropolis()` call; the gradient path is the
right tool since the log-density is smooth.

```{r}
fit_ar1 <- gpu_metropolis(m_ar1$model, data = m_ar1$data,
                          n_iter = 30000, n_chains = 8,
                          method = "mala", warmup = "auto",
                          seed = 1, backend = "cpu")
gpum_diagnose(fit_ar1, plot = FALSE)
```

The classical reference for this exact conditional likelihood is
`arima()`, and the Bernstein-von Mises theorem above says the posterior
mean and standard deviation must land on the maximum likelihood estimate
and its standard error. They do:

```{r}
ml <- arima(y, order = c(1, 0, 0), method = "ML")
comp <- rbind(
  posterior = c(mean(fit_ar1$draws[, , "phi"]),
                sd(fit_ar1$draws[, , "phi"]),
                mean(fit_ar1$draws[, , "mu"])),
  arima = c(coef(ml)["ar1"], sqrt(vcov(ml)["ar1", "ar1"]),
            coef(ml)["intercept"])
)
colnames(comp) <- c("phi", "se(phi)", "mu")
round(comp, 4)
```

The joint posterior view and the information-bound reference read as in
the i.i.d. case studies; the `gpum_crlb()` documentation states why the
observed information of the conditional factorisation is the right
curvature here (the martingale central limit theorem stands in for
Lindeberg-Feller).

```{r}
gpum_pairs(fit_ar1)
```

## 3. Regime 3: the dam as a covariate, dependence plus heterogeneity

The AR(1) treats the 1898 level shift as persistence it has to absorb.
The honest model gives the shift its own parameter: an exogenous indicator
`dam` that is 0 through 1898 and 1 after, entering the conditional mean of
each row. The rows are now neither independent (the lag is still there)
nor identically distributed (the mean moves with the covariate): the
fully non-i.i.d. regime, and still one formula.

```{r}
dam <- as.numeric(year >= 1899)
m_dam <- gpum_ts_model(
  ~ -0.5 * ((y - mu - b * dam - phi * (y_lag1 - mu - b * dam_lag)) /
              exp(ls))^2 - ls - 0.9189385332046727,
  params = c("mu", "b", "phi", "ls"),
  series = list(y = y), order = 1,
  covariates = list(dam = dam, dam_lag = c(0, dam[-length(dam)])),
  prior = ~ -0.5 * ((mu - 1000) / 500)^2 - 0.5 * (b / 500)^2 -
    0.5 * (phi / 2)^2 - 0.5 * ((ls - 5) / 3)^2
)
fit_dam <- gpu_metropolis(m_dam$model, data = m_dam$data,
                          n_iter = 30000, n_chains = 8,
                          method = "mala", warmup = "auto",
                          seed = 1, backend = "cpu")
gpum_diagnose(fit_dam, plot = FALSE)
```

Two questions, answered from the draws with the estimation-based tools of
the Bayesian layer. Did the dam lower the flow, and by how much? And how
much persistence remains once the shift is modelled?

```{r}
gpum_hypothesis(fit_dam, "b", lower = -Inf, upper = 0)
hdi(as.vector(fit_dam$draws[, , "b"]))
round(c(phi_ar1 = mean(fit_ar1$draws[, , "phi"]),
        phi_dam = mean(fit_dam$draws[, , "phi"])), 3)
```

The persistence drops once the level shift is explained: a large part of
what the plain AR(1) read as correlation between consecutive years was
the two levels of the series. This is the classic lesson of confounded
dependence and heterogeneity, visible in four lines.

## 4. Model comparison that respects time

WAIC and PSIS-LOO ask a leave-one-out question, and leaving out a past
observation that later rows condition on is not a forecasting question:
their pointwise exchangeability assumption fails on dependent rows. The
tool that respects the arrow of time is leave-future-out cross-validation
(Burkner, Gabry and Vehtari 2020): refit on the first `t - 1` rows, score
the one-step-ahead predictive density of row `t`, slide forward.
`gpum_lfo()` does this exactly, refitting at every step; the sampler is
fast enough that no importance-sampling approximation is needed.

```{r}
lfo_ar1 <- gpum_lfo(m_ar1, L = 79L, n_iter = 4000, n_chains = 4,
                    seed = 1, backend = "cpu")
lfo_dam <- gpum_lfo(m_dam, L = 79L, n_iter = 4000, n_chains = 4,
                    seed = 1, backend = "cpu")
lfo_ar1
lfo_dam
```

The dam model forecasts the held-out final fifth of the series better,
which is the verdict the level shift already announced. Both formulas
carry their Gaussian normalising constants, so the elpd values are
comparable.

## 5. Observed against generated, in time

The posterior predictive check for a series compares the observed data
with trajectories simulated from the posterior: draw parameters, then
simulate the AR recursion forward from the first observation. The
distribution overlay is the same discipline as in the i.i.d. case
studies, `gpum_density_compare()` on the pooled values, plus a trajectory
band in time.

```{r}
set.seed(2)
draws <- matrix(fit_dam$draws, ncol = fit_dam$n_params)
colnames(draws) <- m_dam$model$params
sim_one <- function(th) {
  ys <- numeric(length(y)); ys[1] <- y[1]
  for (t in 2:length(y)) {
    mt <- th["mu"] + th["b"] * dam[t]
    mtl <- th["mu"] + th["b"] * dam[t - 1]
    ys[t] <- mt + th["phi"] * (ys[t - 1] - mtl) + rnorm(1, 0, exp(th["ls"]))
  }
  ys
}
sims <- sapply(sample(nrow(draws), 200), function(i) sim_one(draws[i, ]))
gpum_density_compare(y, as.vector(sims),
                     main = "Nile flow: observed vs posterior-simulated")
```

```{r}
qs <- apply(sims, 1, quantile, probs = c(0.05, 0.5, 0.95))
plot(year, y, type = "l", lwd = 1.5, ylim = range(qs, y),
     main = "Observed series inside the posterior predictive band",
     xlab = "year", ylab = "flow")
polygon(c(year, rev(year)), c(qs[1, ], rev(qs[3, ])),
        col = grDevices::adjustcolor("steelblue", 0.25), border = NA)
lines(year, qs[2, ], col = "steelblue", lwd = 1.5, lty = 2)
lines(year, y, lwd = 1.5)
abline(v = 1898, lty = 3)
legend("topright", bty = "n", lwd = c(1.5, 1.5), lty = c(1, 2),
       col = c("black", "steelblue"),
       legend = c("observed", "posterior median (90% band)"))
```

## 6. Where this stops, honestly

The conditional-factorisation pattern covers Markov dependence of any
finite order, with or without exogenous covariates, and everything
independent. It does not cover models whose likelihood needs a latent
recursion (GARCH variances, state-space filters) or a fully coupled
covariance (Gaussian processes, `y ~ N(mu, K(theta))`), because the
per-row DSL has no recursion and no matrix operations. For those, Stan's
transformed parameters, sequential Monte Carlo, or INLA are the right
tools, and the roadmap's matrix-DSL tier is where this package would meet
them.

## References

- Billingsley, P. (1961). *Statistical Inference for Markov Processes*.
  University of Chicago Press.
- Borwanker, J., Kallianpur, G., Prakasa Rao, B. L. S. (1971). The
  Bernstein-von Mises theorem for Markov processes. *Annals of
  Mathematical Statistics*, 42(4), 1241-1253.
- Burkner, P.-C., Gabry, J., Vehtari, A. (2020). Approximate
  leave-future-out cross-validation for Bayesian time series models.
  *Journal of Statistical Computation and Simulation*, 90(14), 2499-2523.
- Hall, P., Heyde, C. C. (1980). *Martingale Limit Theory and Its
  Application*. Academic Press.
- van der Vaart, A. W. (1998). *Asymptotic Statistics*. Cambridge
  University Press.
