---
title: "Getting started with gpumetropolis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gpumetropolis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Cap the native CPU backend's Rayon thread pool at two threads while the
# vignette is built: R CMD check farms allow at most two cores.
Sys.setenv(RAYON_NUM_THREADS = "2")
```

`gpumetropolis` is a generic random-walk Metropolis-Hastings sampler. A model
is declared by writing its log-likelihood as an ordinary R formula; the
package compiles that formula to a portable kernel that runs on the CPU and on
the GPU from one source. The sampler advances many independent chains in one
batched pass.

This vignette walks through the workflow on the CPU backend, which every
build provides. The CUDA and Vulkan backends are optional and are noted where
they apply.

```{r library}
library(gpumetropolis)
```

## Declaring a model

A model is declared with `gpum_model()`. The argument `loglik` is a one-sided
formula giving the per-observation log-likelihood, up to an additive constant,
in the parameter and data names. The example below is the Gaussian mean with
a known standard deviation of 2 and a flat prior on the mean. Its posterior is
available in closed form, which makes it a clean check that the sampler
recovers a known quantity.

```{r model}
model <- gpum_model(
  loglik = ~ -((y - mu)^2) / 8,   # sigma = 2, so 0.5 / sigma^2 = 1 / 8
  params = "mu",
  data = "y"
)
model
```

The operations allowed in a formula are `+`, `-`, `*`, `/`, `^`, unary `-`,
and `exp`, `log`, `sqrt`. A symbol that is neither a declared parameter nor a
data column, or a function outside this set, is rejected at compile time with
a clear message.

## Running the sampler

`gpu_metropolis()` runs the sampler. The data is passed as a named list, one
entry per declared data column. `proposal_sd` is the standard deviation of the
Gaussian random-walk proposal and should be tuned to the scale of the
posterior; here the posterior standard deviation is about
`2 / sqrt(n)`.

```{r run}
set.seed(1)
y <- rnorm(5000, mean = 3.4, sd = 2)

fit <- gpu_metropolis(
  model, data = list(y = y),
  proposal_sd = 0.06, n_iter = 2000, n_chains = 8,
  seed = 1, backend = "cpu"
)
fit
```

The printed summary reports the acceptance rate and the posterior mean and
standard deviation of each parameter. The first half of the iterations is
discarded as warmup before `fit$draws` is returned; setting `warmup = 0`
keeps every iteration for inspecting the burn-in trajectory.

## The fit object

The draws are in `fit$draws`, an array of `n_iter - warmup` by `n_chains` by
`n_params`, with the parameters named on the third dimension. The raw count,
the warmup count and the kept count are recorded as `fit$n_iter_total`,
`fit$warmup` and `fit$n_iter`.

```{r draws}
dim(fit$draws)
mu_draws <- fit$draws[, , "mu"]   # iterations (post-warmup) by chains
```

The closed-form posterior of this model is available through
`gaussian_mean_posterior()`, so the sampler can be checked against the exact
answer.

```{r truth}
truth <- gaussian_mean_posterior(y, sigma = 2)
c(sampler_mean = mean(mu_draws), exact_mean = truth$mean)
c(sampler_sd = sd(as.vector(mu_draws)), exact_sd = truth$sd)
```

## Convergence diagnostics

The package computes the split R-hat statistic and the effective sample size.
Both accept the post-warmup `n_iter` by `n_chains` matrix of one parameter's
draws. When the input is already post-warmup, pass `warmup = 0` to keep all
the rows for the diagnostic.

```{r diagnostics}
rhat(mu_draws, warmup = 0)   # split potential scale reduction factor, near 1 at convergence
ess(mu_draws, warmup = 0)    # effective sample size, summed over chains
```

R-hat near 1 indicates the chains have mixed; the effective sample size
reports how many independent draws the autocorrelated output is worth.

## One source, several backends

The same model declaration runs on the CPU, CUDA and Vulkan backends. Only the
`backend` argument changes. The CUDA and Vulkan backends are optional Cargo
features; the lines below run when the package was built with them.

```{r gpu, eval = FALSE}
# the same declaration, dispatched to the GPU
fit_gpu <- gpu_metropolis(
  model, data = list(y = y),
  proposal_sd = 0.06, n_iter = 2000, n_chains = 8,
  seed = 1, backend = "cuda"
)
```

Equivalence between two MCMC runs is distributional, never bit-exact, because
the algorithm is stochastic. `ks_equivalence()` runs a two-sample
Kolmogorov-Smirnov test, thinning the pooled draws to the effective sample
size first, since that test assumes independent draws. The check below
compares two CPU runs from different seeds; the same call compares a CPU run
against a GPU run.

```{r equivalence}
fit_b <- gpu_metropolis(
  model, data = list(y = y),
  proposal_sd = 0.06, n_iter = 2000, n_chains = 8,
  seed = 2, backend = "cpu"
)
ks_equivalence(fit$draws[, , "mu"], fit_b$draws[, , "mu"])
```

## A model with more than one parameter

The DSL is not restricted to one parameter or to the Gaussian mean. The model
below has an unknown mean `mu` and an unknown log standard deviation `ls`;
`proposal_sd` then takes one value per parameter.

```{r two-params}
set.seed(2)
y2 <- rnorm(3000, mean = 2, sd = 1.5)

model2 <- gpum_model(
  loglik = ~ -((y - mu)^2) / (2 * exp(2 * ls)) - ls,
  params = c("mu", "ls"),
  data = "y"
)

init <- cbind(seq(-1, 5, length.out = 6), seq(-1, 1, length.out = 6))
fit2 <- gpu_metropolis(
  model2, data = list(y = y2), init = init,
  proposal_sd = c(0.03, 0.02), n_iter = 4000, seed = 3, backend = "cpu"
)

post2 <- fit2$draws[2001:4000, , ]
c(mu = mean(post2[, , "mu"]), ls = mean(post2[, , "ls"]), log_sd_true = log(1.5))
```

## When the GPU helps

A GPU does not accelerate every MCMC run. The sequential dependence inside one
chain cannot be parallelised. The parallelism comes from two axes: many
independent chains, and the data-parallel evaluation of the log-density. A GPU
pays off when the log-density is expensive to evaluate, that is over a large
data set, or when thousands of chains are run. For a small model with few
chains the transfer overhead dominates and the CPU is faster. The `"auto"`
backend encodes this rule: it selects the CPU for few chains and a GPU backend
for many. The package benchmark in the README characterises the regime in a
refutable way.
